Fast and exact visibility on digitized shapes and application to saliency-aware normal estimation

### 2025-04-22 DGMM2025/paper.tex
fixing some typographic issues

%
\begin{document}
%
    \title{Fast and exact visibility on digitized shapes and application to saliency-aware normal estimation}
%
    \titlerunning{Fast and exact visibility on digitized shapes}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
    \author{Romain Negro\inst{1}\orcidID{0000-XXXX-YYYY-ZZZZ} \and
    Jacques-Olivier Lachaud\inst{1}\orcidID{0000-0003-4236-2133}}
%
    \authorrunning{R. Negro and J.-O. Lachaud}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
    \institute{Universit\'e Savoie Mont Blanc, CNRS, LAMA, F-73000 Chambéry, France\\
    \email{\{romain.negro|jacques-olivier.lachaud\}@univ-smb.fr}}
%
    \maketitle              % typeset the header of the contribution
%
    \begin{abstract}
        Computing visibility on a geometric object requires heavy
        computations since it requires to identify pair of points that
        are visible to each other, i.e. there is a straight segment
        joining them that stays in the close vicinity of the object
        boundary. We propose to exploit a specific representation of
        digital sets based on lists of integral intervals in order to
        compute efficiently the complete visibility graph between
        lattice points of the digital shape. As a quite direct
        application, we show then how we can use visibility to estimate
        the normal vector field of a digital shape in an accurate and
        convergent manner while staying aware of the salient and sharp features of
        the shape.

@@

    Visibility is a fundamental concept in computational geometry and
    digital topology, with applications ranging from computer vision
    (identifying occlusions) to geometric modeling (feature detection, geodesics)
    and computer graphics (path tracing, shadowmap). Visibility within
    polygons in the plane has been extensively studied in the
    litterature \cite{ghosh:2007-book} and is still studied in higher
    dimensional spaces \cite{orourke:2017-book}. The general problem
    has a high complexity. For instance, the 3d visibility complex
    between $n$ spheres and occluding spheres has a complexity
    $O(n^4)$ \cite{durand:2002-tog}. Exact 3d visibility between
    polygons and occluding polygons is often casted into 5d Euclidean
    space derived from Plücker space \cite{nirenstein:2002-ewr}, which
    nicely represents 3d lines and their respective positions. The
    computational cost remains high even with decomposition into
    coarse cells for speed up (several days of computation for a
    million triangles at the time).

    To increase efficiency, several authors have proposed to cast the
    visibility problem into a digital space, generally $\Z^3$ or its
    decomposition into cubical cells. Visibility is simplified as the
    problem if there is a connected digital straight line joining a
    pair of cells without occluding cell(s) (e.g. Soille
    \cite{soille:1994-prl} uses Bresenham line segments). Coeurjolly
    \emph{et al.} \cite{coeurjolly:2004-prl} uses a non-symmetric
    visibility definition and preimage computations to speed up that
    kind of algorithms. Chica \cite{chica:2008-spm} uses a half-cell
    erosion of the complementary space and Bresenham straight lines to
    decide visibility.

    We propose an algorithm that solves the following visibility
    problem: given a digital set $X \subset \Z^d$, two points $p,q$ of
    $X$ are \emph{visible} whenever the Euclidean straight line
    segment $\lbrack p,q \rbrack$ is never at chessboard distance
    ($\infty$-distance) greater or equal to $1$ from $X$. It is
    equivalent to the \emph{cotangency} of $p$ and $q$ in $X$, as
    formalized in \cite{lachaud:2021-dgmm,lachaud:2022-jmiv}, and can
    be viewed as an inclusion problem between sets of lattice points
    in this form. Indeed cells surounding $X$ can be encoded as
    lattice points (Khalimsky coding), as well as the cells covering
    lattice directions. The originality of our method is to encode
    subsets of lattice points as sets of integer intervals and to
    reduce all the inclusion problems as intersections and
    translations of integer intervals. For a given lattice vector
    $\mathbf{t}$, we solve the visibility of every pair of points
    $(p,p+\mathbf{t})$ of $X$. Global visibility is achieved by
    testing all sought directions within a given range, and is
    parallelized straightforwardly.

    We focus on solving efficiently this problem because it is crucial
    in computing geometric normals along digital surfaces that are
    aware of salient features. Indeed, 2d discrete tangent estimators
    based on tangency or discrete line segments have proven to be both
    sensitive to sharp corners while multigrid convergent along the
    boundary of digitized smooth shapes
    \cite{feschet:1999-dgci,lachaud:2007-ivc,nguyen:2011-pr}. Many ad
    hoc methods have been proposed for tangent plane or normal
    estimation on 3d digital surfaces
    (e.g. \cite{fourey:2009-cg,charrier:2011-iwcia,Cuel:2014-dgci,Lachaud:2017-lnm,mareche:2024-ispr}).
    If two of these methods \cite{Cuel:2014-dgci,Lachaud:2017-lnm}
    have been shown to be multigrid convergent on digitization of
    shapes with smooth boundary, their formulation intrinsically
    smoothes features, e.g. in contrast with the method of
    Mar{\^e}ch{\'e} \emph{et al.}  \cite{mareche:2024-ispr}, whose
    convergence is not established. We propose here a normal estimator
    defined as the most orthogonal vector to the cone of visibility at
    the point of interest. Visibility is high in smooth regions and
    the induced visibility vectors are good approximations of tangent
    vectors. On the contrary the visibility is blocked at sharp
    features, so the normal vector is then not influenced by the
    geometry on the other side of the features. We demonstrate the
    superiority of this normal estimator for estimating curvatures
    along digital surfaces with sharp features.

    Note that our algorithm for computing visibility is in fact much
    more general. The same algorithm could be used to recognize if any
    pattern of lattice points is present on a shape. It is similar to
    an erosion algorithm where the structuring element is the chosen
    pattern, and could also be used for morphological operations
    \cite{soille1999morphological}.  The paper outline is as
    follows. In Section~2, we recall some definitions and present
    tangency of chords and its equivalence to visibility.  Section~3
    describes the specific data structure for encoding set of grid
    cells, and the main algorithm that exploits this encoding to
    compute visibility exactly.  Section~4 presents experimental
    results on visibility, its usage as a discrete normal estimator,
    and how it improves curvature estimates on digital surfaces.
    Finally, Section 5 concludes and gives some perspectives to this
    work, notably other usages of this algorithm in the field.

>>>>

%
\begin{document}
%
    \title{Fast and exact visibility on digitized shapes and application to saliency-aware normal estimation}
%
    \titlerunning{Fast and exact visibility on digitized shapes}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
    \author{Romain Negro\inst{1}\orcidID{0000-XXXX-YYYY-ZZZZ} \and
    Jacques-Olivier Lachaud\inst{1}\orcidID{0000-0003-4236-2133}}
%
    \authorrunning{R. Negro and J.-O. Lachaud}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
    \institute{Universit\'e Savoie Mont Blanc, CNRS, LAMA, F-73000 Chambéry, France\\
    \email{\{romain.negro|jacques-olivier.lachaud\}@univ-smb.fr}}
%
    \maketitle              % typeset the header of the contribution
%
    \begin{abstract}
        Computing visibility on a geometric object requires heavy
        computations since it requires to identify pairs of points that
        are visible to each other, i.e.\ there is a straight segment
        joining them that stays in the close vicinity of the object
        boundary. We propose to exploit a specific representation of
        digital sets based on lists of integral intervals in order to
        compute efficiently the complete visibility graph between
        lattice points of the digital shape. As a quite direct
        application, we show then how we can use visibility to estimate
        the normal vector field of a digital shape in an accurate and
        convergent manner while staying aware of the salient and sharp features of
        the shape.

@@

    Visibility is a fundamental concept in computational geometry and
    digital topology, with applications ranging from computer vision
    (identifying occlusions) to geometric modeling (feature detection, geodesics)
    and computer graphics (path tracing, shadowmap). Visibility within
    polygons in the plane has been extensively studied in the
    litterature~\cite{ghosh:2007-book} and is still studied in higher
    dimensional spaces~\cite{orourke:2017-book}. The general problem
    has a high complexity. For instance, the 3d visibility complex
    between $n$ spheres and occluding spheres has a complexity
    $O(n^4)$~\cite{durand:2002-tog}. Exact 3d visibility between
    polygons and occluding polygons is often casted into 5d Euclidean
    space derived from Plücker space~\cite{nirenstein:2002-ewr}, which
    nicely represents 3d lines and their respective positions. The
    computational cost remains high even with decomposition into
    coarse cells for speed up (several days of computation for a
    million triangles at the time).

    To increase efficiency, several authors have proposed to cast the
    visibility problem into a digital space, generally $\Z^3$ or its
    decomposition into cubical cells. Visibility is simplified as the
    problem if there is a connected digital straight line joining a
    pair of cells without occluding cell(s) (e.g.\ Soille~\cite{soille:1994-prl}
    uses Bresenham line segments). Coeurjolly \emph{et al.}~\cite{coeurjolly:2004-prl}
    uses a non-symmetric visibility definition and preimage computations
    to speed up that kind of algorithms. Chica~\cite{chica:2008-spm}
    uses a half-cell erosion of the complementary space and Bresenham
    straight lines to decide visibility.

    We propose an algorithm that solves the following visibility
    problem: given a digital set $X \subset \Z^d$, two points $p,q$ of
    $X$ are \emph{visible} whenever the Euclidean straight line
    segment $\lbrack p,q \rbrack$ is never at chessboard distance
    ($\infty$-distance) greater or equal to $1$ from $X$. It is
    equivalent to the \emph{cotangency} of $p$ and $q$ in $X$, as
    formalized in~\cite{lachaud:2021-dgmm,lachaud:2022-jmiv}, and can
    be viewed as an inclusion problem between sets of lattice points
    in this form. Indeed cells surounding $X$ can be encoded as
    lattice points (Khalimsky coding), as well as the cells covering
    lattice directions. The originality of our method is to encode
    subsets of lattice points as sets of integer intervals and to
    reduce all the inclusion problems as intersections and
    translations of integer intervals. For a given lattice vector
    $\mathbf{t}$, we solve the visibility of every pair of points
    $(p,p+\mathbf{t})$ of $X$. Global visibility is achieved by
    testing all sought directions within a given range, and is
    parallelized straightforwardly.

    We focus on solving efficiently this problem because it is crucial
    in computing geometric normals along digital surfaces that are
    aware of salient features. Indeed, 2d discrete tangent estimators
    based on tangency or discrete line segments have proven to be both
    sensitive to sharp corners while multigrid convergent along the
    boundary of digitized smooth shapes~\cite{feschet:1999-dgci,lachaud:2007-ivc,nguyen:2011-pr}.
    Many ad hoc methods have been proposed for tangent plane or normal
    estimation on 3d digital surfaces
    (e.g. \cite{fourey:2009-cg,charrier:2011-iwcia,Cuel:2014-dgci,Lachaud:2017-lnm,mareche:2024-ispr}).
    If two of these methods~\cite{Cuel:2014-dgci,Lachaud:2017-lnm}
    have been shown to be multigrid convergent on digitization of
    shapes with smooth boundary, their formulation intrinsically
    smoothes features, e.g.\ in contrast with the method of
    Mar{ê}ch{é} \emph{et al.}~\cite{mareche:2024-ispr}, whose
    convergence is not established. We propose here a normal estimator
    defined as the most orthogonal vector to the cone of visibility at
    the point of interest. Visibility is high in smooth regions and
    the induced visibility vectors are good approximations of tangent
    vectors. On the contrary the visibility is blocked at sharp
    features, so the normal vector is then not influenced by the
    geometry on the other side of the features. We demonstrate the
    superiority of this normal estimator for estimating curvatures
    along digital surfaces with sharp features.

    Note that our algorithm for computing visibility is in fact much
    more general. The same algorithm could be used to recognize if any
    pattern of lattice points is present on a shape. It is similar to
    an erosion algorithm where the structuring element is the chosen
    pattern, and could also be used for morphological operations~\cite{soille1999morphological}.
    The paper outline is as follows. In Section~2, we recall some
    definitions and present tangency of chords and its equivalence
    to visibility. Section~3 describes the specific data structure
    for encoding set of grid cells, and the main algorithm that
    exploits this encoding to compute visibility exactly. Section~4
    presents experimental results on visibility, its usage as a
    discrete normal estimator, and how it improves curvature estimates
    on digital surfaces. Finally, Section 5 concludes and gives some
    perspectives to this work, notably other usages of this algorithm
    in the field.
